Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-04-24 WholePlotTrajectories.Rmd
Discussion

The monitoring of disturbed forest communities confirmed a limited negative impact of disturbance on species richness, as observed on several post logging surveys [@Cannon1998;@Baraloto2012a].
Thirty years after disturbance, the most disturbed plots had recovered their initial genus richness, similar to this of control plots.
After low intensity plots the genus richness substantially increased for low intensity plots, reaching a gain of almost 12 genuses for some plots.

The evenness of all disturbed plots and the richness of low disturbance plots on the other hand mainly followed an asymptotic growth sharply increasing until 15 years after disturbance.
Such increased evenness is explained by a much more homogeneous species distribution after disturbance.
Communities composition after disturbance is either due to the old, pre-disturbance survivors or to the recruited trees.
The composition of old survivors proved to mirror the initial community [@Herault2018], the observed composition turnover would stem from an enhanced the growth and survival of previously infrequent species which, along with the IDH, reorganizes the typical high dominance structure of hyperdiverse mature forests.
This increase in taxonomic diversity was accompanied by an increased taxonomic dissimilarity to initial state along time, translating a a taxonomic turnover benefiting to pioneers and light demanding species, as illustrated by the communities functional shifts towards resource-acquisitive strategies (sharp increase in the SLA, leaf thickness and bark thickness and decrease in wood density, leaf toughness and maximum height) [@Westoby1998;@Wright2004;@Reich2014].
Post-disturbance changes in abiotic environment and competitive pressure then well favored pioneers outcompeting other species when abiotic resources are abundant but otherwise excluded in mature forests by long-lived, resistant and shade tolerant species.
The trajectories of proportions among seed mass classes besides revealed the enhanced growth of small seeded, large dispersive, species and therefore the importance of species reproductive strategy which further supported the role of species dispersion and demographic strategy for the post-disturbance dynamics, as suggested by the IDH [@TerSteege2001;@Flores2006;@Haddad2008].

@@

The functional diversity followed a hump-back trajectory which maximum was positively correlated to the disturbance intensity and after 30 years, all disturbed plots had recovered diversities close to their initial values. 
Similar hump-back trajectories were followed by all leaf and stem functional traits and life history traits, to the exception of leaf chlorophyll content.
Although the return to initial values was still ongoing the recovery of functional diversity and average communities trait values was undeniable.
This translated the recovery of ecosystems processes [@Guariguata2001], as functional traits are the most direct link between biodiversity and ecosystem functioning [@Diaz2005].
Same recovery trajectories were followed by taxonomic and functional composition which argued for communities convergence and the maintenance of plots initial difference [@Hubbell1999;@Molino2001;@Baraloto2012a].
These compositon difference, and probably also some abiotic parameters, besides strongly determined the maximum and time path of the trajectories which, despite an homogeneous trend  were quite variable among plots.

@@

Despite the consistent recovery of initial diversity and composition there was a time lag between the communities taxonomic and functional characteristics.
Taxonomic composition had longer time path and while communities had recovered their functional diveristy, their taxonomic composition and evenness remained higher than before disturbance.
This delay between functional and taxonomic dynamics was already observed for grasslands [@Tilman1997; @Mouillot2011] and more recently for tropical forests [@Lohbeck2015;@Guariguata2001].
Communities functional diversity rely, according to the "vegetation quantity effect" [@Grime1998], on the functional strategy of dominant species.
Communities functional trajectory is then first driven by the increase of dominant species diversity and evenness following disturbance and then by the emergence of recruitment of species progressively resembling the old pre-disturbance community which decreased the diversity.
Dominant species then restored the functional diversity and the dominant functional type of the community although the functional and taxonomic composition remained altered and the infrequent species of initial communities are still missing to plots recovery.
The functional traits trajectories suggest that those species would match the dominant functional type therefore increasing communities functional redundancy.
The functional overlap between species is typical of the huge biodiversity of tropical forests [@Bellwood2006], and proved far from recovery 03 years after disturbance.
It is though a determinant of forests resilience and should be accounted for [@Trenbath1999;@Elmqvist2003;@Diaz2005].

Long-term alteration of functional redundancy besides came along with persistent compositional changes which most probably impact disturbance resistant species [@Haddad2008] and probably lianas or epiphytes [@Martin2013], and the modification of environmental conditions like soils nutrient cycling and compaction [@Olander2005] forest resilience remains under question [@Chazdon2003a].
New conditions would not only be longer lasting but self-maintained as tied to disturbance regime [@Burslem2000].
Specificaly, this would impair species contingent to undisturbed forests, threatening their maintenance, and run the risk to loose cornerstone species and trigger unexpected ecological consequences [@Jones1994;@Diaz2005;@Gardner2007].

@@

Our study showed the significant impact of disturbance on tropical forests communities.
The subsequent diversity trajectories confirmed the intermediate disturbance hypothesis debated for tropical forests through their correlation with disturbance intensity.
Besides it revealed the contrasting response of taxonomic and functional characteristics, specifically the decoupling between communities taxonomic evenness and their functional diversity and dominant functional traits values.
The long-term disturbance trajectories observed highlighted the unachieved but consistent recovery of communities assembly for the lowest disturbance intensity but questioned it after higher disturbance.

>>>>

The monitoring of disturbed forest communities confirmed a limited negative impact of disturbance on species richness, as observed on several post logging surveys [@Cannon1998;@Baraloto2012a].
Thirty years after disturbance, the genus richness was restored to initial plot values after high disturbance intensity while it substantially increased after low disturbance intensity, reaching a gain of almost 12 genuses for some plots.

Both richness and evenness after disturbance followed asymptotic trajectories, sharply increasing for 15 years before stabilizing at higher values than those of control plots.
An increase of communities evenness stems from a higher homogeneity of species distribution.
After disturbance, communities are made of the old, pre-disturbance survivors and the newly recruited trees: changes in composition and abundance distribution are to be found in the recruitment processes and in the pre-disturbance survivors mortality.
The composition of old survivors proved to mirror the initial community [@Herault2018], which leads to support a species turnover among recruited trees with enhanced growth and survival of previously infrequent species.
Indeed, the increase in taxonomic diversity was accompanied by an increase of taxonomic dissimilarity with plots initial state and a functional shift towards resource-acquisitive strategies (sharp increase in the SLA, leaf thickness and bark thickness and decrease in wood density, leaf toughness and maximum height) [@Westoby1998;@Wright2004;@Reich2014].
Disturbance then causes a reorganization of the typical high dominance structure of hyperdiverse mature forests after disturbance, benefiting to pioneers and light demanding species. 
Likely, the changes in abiotic environment and competitive pressure favored pioneers which outcompete others in non limiting resources but are excluded in mature forests by long-lived, more resistant and shade tolerant species.

As stated by the IDH, communities dynamics after disturbance then rely on species functional strategy and corresponding ability to fill the environmental niches made available by disturbance.
Recruited species then mix with pre-disturbance ones from which they differ and constitue a community all the more diversified that the disturbance is intense [@Molino2001].

@@

The trajectories of almost all leaf and stem functional traits, life history traits,and functional diversity followed a hump-back trajectory which maximum were positively correlated to the disturbance intensity, and were recovering towards initial values after 30 years that trajectory. 
Functoinal traits and communities functional diversity are the most direct link between biodiversity and ecosystem functioning [@Diaz2005]: these trajectories reflected a consistent recovery in the long term of ecosystem processes [@Guariguata2001].

Same hump-back trajectories were followed by the composition of plots species and functoinal composition.
This similarly meant a recovery of the pre-disturbance state and the maintenance of the initial differences among plots which supports the idea of a convergence of communities composition [@Hubbell1999;@Molino2001;@Baraloto2012a].
Although plots followed homogeneous trends these initial differences in compositon difference, and probably also some abiotic parameters, strongly determined the maximum and time path of the trajectories [@Anderson2007]•.

@@

Despite the consistent recovery of initial diversity and composition there was a time lag between the trajectories followed by taxonomic and functional characteristics.
Taxonomic composition had longer time path and while communities had recovered their functional characteristics, their taxonomic composition and evenness remained altered by disturbance.
This delay between functional and taxonomic dynamics was already observed for grasslands [@Tilman1997; @Mouillot2011] and more recently for tropical forests [@Lohbeck2015;@Guariguata2001].
The functional diversity of communities rely on the functional strategy of dominant species, according to the "vegetation quantity effect" [@Grime1998].
The functional trajectories of communities are then driven first by the increase of diversity and evenness of dominant species and then by the recruitment of species resembling the old pre-disturbance community which reduced the overall diversity.
Although dominant species restoredthe functional pre-disturbance characteristics, communities composition remained altered and the infrequent species of initial communities remain missing to achieve plots recovery.
According to the funcitonal traits trajectories, those infrequent species would match the dominant functional type and would participate to the functional redundancy.
The functional redundancy, thas is the functional overlap between species that is typical of the huge biodiversity of tropical forests [@Bellwood2006], therefore was not restored 30 years after disturbance and this is majr to consider as it defines forests' resilience [@Trenbath1999;@Elmqvist2003;@Diaz2005].

Besides the long-term alteration of functional redundancy, the persistent compositional changes which might have favored disturbance resistant species [@Haddad2008], lianas or epiphytes [@Martin2013] and would probably have come along with environmental change, like in the soils nutrient cycling and compaction [@Olander2005].
These presistent changes highly question forest's resilience[@Chazdon2003a].
New conditions would not only be longer lasting but self-maintained as tied to disturbance regime [@Burslem2000].
Specificaly, this would impair species contingent to undisturbed forests, threatening their maintenance, and run the risk to loose cornerstone species and trigger unexpected ecological consequences [@Jones1994;@Diaz2005;@Gardner2007].

@@

Our study showed the significant impact of disturbance on tropical forests communities.
It validated the consistency of the IDH in the long term and revealed the contrasting response of taxonomic and functional characteristics with the decoupling between communities taxonomic evenness and their functional diversity and th dominant functional strategies.
Communities recovery remained unachieved but consistent recovery of communities assembly for the lowest disturbance intensity but questioned it after higher disturbance.
